Trial by Fire: Diachronic Regional Sentiment Towards American Wildfires

### 2022-12-12 paper.Rmd
final draft

One of North America's most noticeable symptoms of climate change has been a dramatic increase in the volume and intensity of wildfires in its western regions. August and September were previously defined by clear skies and fair weather, but fire and smoke have become a recurrent factor in this area. According to the National Interagency Fire Center (NIFC), more acreage burned nationally in each of the last seven years than any single year but one in the first ten years this data was collected (1983-1993, ex. 1988).^[See https://www.nifc.gov/fire-information/statistics/wildfires.] With more and more people affected by the increase in wildfire activity, it's reasonable to infer an increasing outcry of negativity towards this devastating symptom of climate change. However, less is known about regional attitudes towards this recent trend. As with many other catastrophes, is it true that wildfires are "out of sight, out of mind" for those not directly affected?

@@

We compared two different sentiment analysis algorithms in the hopes of gaining some insight into changing public perceptions of climate change.

@@

The size of each fire's corpus was unfortunately still far from uniform, possibly due to unequal media coverage, but we found this to be a reasonable compromise among a set of unappealing options. We found both Camp Fire and Dixie Fire generated far more content on Twitter under our listed constraints when compared to Tubbs Fire and the Bay Area Fires. Table 1 shows the volume of tweets per fire.

@@

Filtering for tweets that explicitly mentioned the fire by name significantly limited the total number of results. This proved to be particularly debilitating when working on location-based analyses, since only a subset of that limited population reported their location while tweeting. To alleviate this problem, we chose to use much more generic search terms when conducting our location-based analyses. For the purposes of this paper, we found the search term "fire" tended to return the most relevant results (as compared to similarly broad terms like "smoke" or "burn"), though this was not a systematic decision. Nonetheless, it is hard to argue that "fire" is not an appropriate search term to use when conducting sentiment analysis on wildfires.

@@

To overcome this issue, we decided to use Cardiff's roBERTa-base model for sentiment analysis as well.^[https://huggingface.co/cardiffnlp/twitter-roberta-base-sentiment-latest] This model was trained on roughly 124 million tweets from January 2018 to December 2021, using the TweetEval benchmark to fine tune the sentiment analysis. The model outputs probabilities that a tweet is positive, neutral, or negative. For the purposes of our analysis, we simplified the output to a categorical variable representing the sentiment with the highest probability. (Strength of sentiment was not captured.)

@@

To track to sentiment shift through time, we chronologically ordered the fires (Tubbs Fire, Camp Fire, Bay Area Fires, Dixie Fire) and calculated the proportion of each sentiment for each fire. We used proportions because of the unequal distribution of tweets across the fires. We apply the Bonferroni correction to our 95% confidence intervals to account for multiple testing.

@@

In our first data set (Table 1), sample size varied by more than an order of magnitude between certain fires. It was surprising that we could not find more tweets for the Bay Area and Tubbs fires, given their proximity to San Francisco. We originally suspected that the Camp Fire tweet set contained many false positives, but it appears that fire generated more engagement than expected due to the almost total devastation (and subsequent press coverage) of the towns of Paradise, Concow, Magalia, and Butte Creek Canyon. The Dixie Fire was similarly destructive. As the largest ever single wildfire in the state's history, part of its impact included popular Lassen National Park, the coverage of which likely increased engagement on Twitter.

Given more time, it would have been helpful to correct for fire intensity. These fires were all record-setting, but they took place in different conditions. Some were closer to urban centers, and some were quite far. Some took place in the middle of summer and were coincident with a large set of other wildfires, while others (particularly the Camp Fire) took place later in the season, in near isolation. Early September of 2020, near the peak of the Bay Area fires, was a particularly striking example of how catastrophic these coincident fires could be (see Figure 4). It would be particularly interesting to compare sentiment to air quality, though we suspect the combination of (typically) rapid fluctuations in air quality and the sparsity of location-enabled tweets on any given day would make robust statistical conclusions difficult.

>>>>

One of North America's most noticeable symptoms of climate change has been a dramatic increase in the volume and intensity of wildfires in its western regions. August and September were previously defined by clear skies and fair weather, but fire and smoke have become recurrent factors in this area. Since 2015, more acreage has burned nationally each year compared to any year between 1983 (the first year this data was collected) and 1993, excluding 1988.^[See https://www.nifc.gov/fire-information/statistics/wildfires.] With more and more people affected by this increase in wildfire activity, it's reasonable to infer an increasing outcry of negativity towards this symptom of climate change. However, less is known about regional attitudes towards this recent trend. As with many other catastrophes, is it true that wildfires are "out of sight, out of mind" for those not directly affected?

@@

@@

The size of each fire's corpus was unfortunately still far from uniform, possibly due to unequal media coverage, but we found this to be a reasonable compromise among a set of imperfect options. We found that both the Camp Fire and the Dixie Fire generated far more content on Twitter under our listed constraints when compared to the Tubbs Fire and the Bay Area Fires. Table 1 shows the volume of tweets per fire.

@@

Filtering for tweets that explicitly mentioned the fire by name significantly limited the total number of results. This proved to be particularly restricting when working on location-based analyses, since only a subset of that limited population reported their location while tweeting. To alleviate this problem, we chose to use much more generic search terms when conducting our location-based analyses. For the purposes of this paper, we found the search term "fire" tended to return the most relevant results (as compared to similarly broad terms like "smoke" or "burn"), though this was not a systematic decision. Nonetheless, it is hard to argue that "fire" is not an appropriate search term to use when conducting sentiment analysis on wildfires.

@@

To overcome this issue, we decided to additionally use Cardiff's roBERTa-base model for sentiment analysis.^[https://huggingface.co/cardiffnlp/twitter-roberta-base-sentiment-latest] This model was trained on roughly 124 million tweets from January 2018 to December 2021, using the TweetEval benchmark to fine tune the sentiment analysis. The model outputs probabilities that a tweet is positive, neutral, or negative. For the purposes of our analysis, we simplified the output to a categorical variable representing the sentiment with the highest probability. (Strength of sentiment was not captured.)

@@

To track changes in sentiment over time, we chronologically ordered the fires (Tubbs Fire, Camp Fire, Bay Area Fires, Dixie Fire) and calculated the proportion of each sentiment for each fire. We used proportions because of the unequal distribution of tweets across the fires. Additionally, we applied the Bonferroni correction to our 95% confidence intervals to account for multiple testing.

@@

In our first data set (Table 1), sample size varied by more than an order of magnitude between certain fires. It was surprising that we could not find more tweets for the Bay Area and Tubbs fires, given their proximity to San Francisco. We originally suspected that the Camp Fire tweet set contained many false positives, but it appears that fire generated more engagement than expected due to the almost total devastation (and subsequent press coverage) of the towns of Paradise, Concow, Magalia, and Butte Creek Canyon. The Dixie Fire was similarly destructive. As the largest ever single wildfire in the state's history, part of its impact included popular Lassen National Park, the coverage of which may have increased engagement on Twitter.

Given more time, it would have been helpful to correct for fire intensity. These fires were all record-setting, but they took place in different conditions. Some were closer to urban centers, and some were quite far. Some took place in the middle of summer and were coincident with a large set of other wildfires, while others (particularly the Camp Fire) took place later in the season, in near isolation. Satellite imagery taken near the peak of the Bay Area fires illustrates the extent to which the West Coast was engulfed in smoke, much of which was from other, smaller fires (see Figure 4). It would be particularly interesting to compare sentiment to air quality, though we suspect the combination of (typically) rapid fluctuations in air quality and the sparsity of location-enabled tweets on any given day would make robust statistical conclusions difficult.
